---
title: "Chapter 6. Calibrating Hedonic Pricing Model for Private Highrise Property with GWR Method"
editor: visual
---

# Overview

We will be looking at how to build [hedonic pricing](https://www.investopedia.com/terms/h/hedonicpricing.asp) models by using GWR methods. The dependent variable is the resale prices of condominium in 2015. The independent variables are divided into either structural and locational.

# The Study

**Geographically weighted regression (GWR)** is a spatial statistical technique that takes non-stationary variables into consideration (e.g., climate; demographic factors; physical environment characteristics) and models the local relationships between these independent variables and an outcome of interest (also known as dependent variable).

# Initialization

Initialize the following packages using pacman:

```{r}
pacman::p_load(olsrr, corrplot, ggpubr, sf, spdep, GWmodel, tmap, tidyverse, gtsummary)
```

The R packages needed for this exercise are as follows:

-   R package for building OLS and performing diagnostics tests

    -   [**olsrr**](https://olsrr.rsquaredacademy.com/)

-   R package for calibrating geographical weighted family of models

    -   [**GWmodel**](https://cran.r-project.org/web/packages/GWmodel/)

-   R package for multivariate data visualisation and analysis

    -   [**corrplot**](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html)

-   Spatial data handling

    -   **sf**

-   Attribute data handling

    -   **tidyverse**, especially **readr**, **ggplot2** and **dplyr**

-   Choropleth mapping

    -   **tmap**

# Data Acquisition and Import

## Data Acquisition

-   URA Master Plan subzone boundary in shapefile format (i.e. *MP14_SUBZONE_WEB_PL*)

-   condo_resale_2015 in csv format (i.e. *condo_resale_2015.csv*)

## Importing Data

### Geospatial Data

Using st_read() function of sf package to import the relevant shapefile into R.

```{r}
mpsz = st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
```

```{r}
mpsz_svy21 <- st_transform(mpsz, 3414)
st_crs(mpsz_svy21)
```

After transforming the projection metadata, you can varify the projection of the newly transformed *mpsz_svy21* by using `st_crs()` of **sf** package.

```{r}
st_bbox(mpsz_svy21)
```

### Aspatial Data

```{r}
condo_resale = read_csv("data/aspatial/Condo_resale_2015.csv")
glimpse(condo_resale)
```

```{r}
head(condo_resale$LONGITUDE) #see the data in XCOORD column
```

```{r}
head(condo_resale$LATITUDE) #see the data in YCOORD column
```

```{r}
summary(condo_resale)
```

#### Converting aspatial data frame into a sf object

Currently, the condo_resale tibble data frame is aspatial. We will convert it to a sf object. The code chunk below converts condo_resale data frame into a simple feature data frame by using st_as_sf() of sf packages.

```{r}
condo_resale.sf <- st_as_sf(condo_resale,
                            coords = c("LONGITUDE", "LATITUDE"),
                            crs=4326) %>%
  st_transform(crs=3414)
```

Notice that `st_transform()` of **sf** package is used to convert the coordinates from wgs84 (i.e. crs:4326) to svy21 (i.e. crs=3414).

Next, `head()` is used to list the content of *condo_resale.sf* object.

```{r}
head(condo_resale.sf)
```

# Exploratory Data Analysis (EDA)

## Statistical Graphics

```{r}
ggplot(data=condo_resale.sf, aes(x=`SELLING_PRICE`)) +
  geom_histogram(bins=20, color="black", fill="light blue")
```

The figure above reveals a right skewed distribution. This means that more condominium units were transacted at relative lower prices. Statistically, the skewed dsitribution can be normalised by using log transformation.

Derive a new variable called *LOG_SELLING_PRICE* by using a log transformation on the variable *SELLING_PRICE*. It is performed using `mutate()` of **dplyr** package.

```{r}
condo_resale.sf <- condo_resale.sf %>%
  mutate(`LOG_SELLING_PRICE` = log(SELLING_PRICE))

ggplot(data=condo_resale.sf, aes(x=`LOG_SELLING_PRICE`)) +
  geom_histogram(bins=20, color="black", fill="light blue")
```

Notice that the distribution is relatively less skewed after the transformation.

## Multiple Histogram Plots distribution of variables

```{r}
AREA_SQM <- ggplot(data=condo_resale.sf, aes(x= `AREA_SQM`)) + 
  geom_histogram(bins=20, color="black", fill="light blue")

AGE <- ggplot(data=condo_resale.sf, aes(x= `AGE`)) +
  geom_histogram(bins=20, color="black", fill="light blue")

PROX_CBD <- ggplot(data=condo_resale.sf, aes(x= `PROX_CBD`)) +
  geom_histogram(bins=20, color="black", fill="light blue")

PROX_CHILDCARE <- ggplot(data=condo_resale.sf, aes(x= `PROX_CHILDCARE`)) + 
  geom_histogram(bins=20, color="black", fill="light blue")

PROX_ELDERLYCARE <- ggplot(data=condo_resale.sf, aes(x= `PROX_ELDERLYCARE`)) +
  geom_histogram(bins=20, color="black", fill="light blue")

PROX_URA_GROWTH_AREA <- ggplot(data=condo_resale.sf, 
                               aes(x= `PROX_URA_GROWTH_AREA`)) +
  geom_histogram(bins=20, color="black", fill="light blue")

PROX_HAWKER_MARKET <- ggplot(data=condo_resale.sf, aes(x= `PROX_HAWKER_MARKET`)) +
  geom_histogram(bins=20, color="black", fill="light blue")

PROX_KINDERGARTEN <- ggplot(data=condo_resale.sf, aes(x= `PROX_KINDERGARTEN`)) +
  geom_histogram(bins=20, color="black", fill="light blue")

PROX_MRT <- ggplot(data=condo_resale.sf, aes(x= `PROX_MRT`)) +
  geom_histogram(bins=20, color="black", fill="light blue")

PROX_PARK <- ggplot(data=condo_resale.sf, aes(x= `PROX_PARK`)) +
  geom_histogram(bins=20, color="black", fill="light blue")

PROX_PRIMARY_SCH <- ggplot(data=condo_resale.sf, aes(x= `PROX_PRIMARY_SCH`)) +
  geom_histogram(bins=20, color="black", fill="light blue")

PROX_TOP_PRIMARY_SCH <- ggplot(data=condo_resale.sf, 
                               aes(x= `PROX_TOP_PRIMARY_SCH`)) +
  geom_histogram(bins=20, color="black", fill="light blue")

ggarrange(AREA_SQM, AGE, PROX_CBD, PROX_CHILDCARE, PROX_ELDERLYCARE, 
          PROX_URA_GROWTH_AREA, PROX_HAWKER_MARKET, PROX_KINDERGARTEN, PROX_MRT,
          PROX_PARK, PROX_PRIMARY_SCH, PROX_TOP_PRIMARY_SCH,  
          ncol = 3, nrow = 4)
```

## Drawing Statistical Point Map

We want to reveal the geospatial distribution condominium resale prices in Singapore. The map will be prepared by using **tmap** package.

```{r}
tmap_mode("view")
tm_shape(mpsz_svy21)+
  tm_polygons() +
tm_shape(condo_resale.sf) +  
  tm_dots(col = "SELLING_PRICE",
          alpha = 0.6,
          style="quantile") +
  tm_view(set.zoom.limits = c(11,14))+
  tmap_options(check.and.fix = TRUE)
```

[`tm_dots()`](https://www.rdocumentation.org/packages/tmap/versions/2.2/topics/tm_symbols) is used instead of `tm_bubbles()`. `set.zoom.limits` argument of `tm_view()` sets the minimum and maximum zoom level to 11 and 14 respectively.

```{r}
tmap_mode("plot")
```

`tmap_mode()` will need to change back to plot before continuing.

# Hedonic Pricing Modelling in R

Building hedonic pricing models for condominium resale units using lm() of R base.

## Simple Linear Regression Method

We will build a simple linear regression model by using SELLING_PRICE as the dependent variable and AREA_SQM as the independent variable.

```{r}
condo.slr <- lm(formula=SELLING_PRICE ~ AREA_SQM, data = condo_resale.sf)
```

`lm()` returns an object of class \"lm\" or for multiple responses of class c(\"mlm\", \"lm\").

The functions `summary()` and `anova()` can be used to obtain and print a summary and analysis of variance table of the results. The generic accessor functions coefficients, effects, fitted.values and residuals extract various useful features of the value returned by `lm`.

```{r}
summary(condo.slr)
```

The output report reveals that the SELLING_PRICE can be explained by using the formula:

          *y = -258121.1 + 14719x1*

The R-squared of 0.4518 reveals that the simple regression model built is able to explain about 45% of the resale prices.

Since p-value is much smaller than 0.0001, we will reject the null hypothesis that mean is a good estimator of SELLING_PRICE. This will allow us to infer that simple linear regression model above is a good estimator of *SELLING_PRICE*.

The **Coefficients:** section of the report reveals that the p-values of both the estimates of the Intercept and ARA_SQM are smaller than 0.001. In view of this, the null hypothesis of the B0 and B1 are equal to 0 will be rejected. As a results, we will be able to infer that the B0 and B1 are good parameter estimates.

To visualise the best fit curve on a scatterplot, we can incorporate `lm()` as a method function in ggplot\'s geometry as shown in the code chunk below.

```{r}
ggplot(data=condo_resale.sf,  
       aes(x=`AREA_SQM`, y=`SELLING_PRICE`)) +
  geom_point() +
  geom_smooth(method = lm)
```

## Multiple Linear Regression Method

### Visualizing the relationships of the independent variables

```{r}
corrplot(cor(condo_resale[, 5:23]), diag = FALSE, order = "AOE",
         tl.pos = "td", tl.cex = 0.5, method = "number", type = "upper")
```

It is important to ensure that the indepdent variables used are not highly correlated to each other. If these highly correlated independent variables are used in building a regression model by mistake, the quality of the model will be compromised. This phenomenon is known as **multicollinearity** in statistics.

Correlation matrix is commonly used to visualise the relationships between the independent variables. Beside the `pairs()` of R, there are many packages support the display of a correlation matrix. In this section, the [**corrplot**](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html) package will be used.

```{r}
corrplot(cor(condo_resale[, 5:23]), diag = FALSE, order = "AOE",
         tl.pos = "td", tl.cex = 0.5, method = "number", type = "upper")
```

### Building a hedonic pricing model using multiple linear regression method

The code chunk below using `lm()` to calibrate the multiple linear regression model.

```{r}
condo.mlr <- lm(formula = SELLING_PRICE ~ AREA_SQM + AGE    + 
                  PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE +
                  PROX_URA_GROWTH_AREA + PROX_HAWKER_MARKET + PROX_KINDERGARTEN + 
                  PROX_MRT  + PROX_PARK + PROX_PRIMARY_SCH + 
                  PROX_TOP_PRIMARY_SCH + PROX_SHOPPING_MALL + PROX_SUPERMARKET + 
                  PROX_BUS_STOP + NO_Of_UNITS + FAMILY_FRIENDLY + FREEHOLD, 
                data=condo_resale.sf)
summary(condo.mlr)
```

### Preparing Publication Quality Table: olsrr method

```{r}
condo.mlr1 <- lm(formula = SELLING_PRICE ~ AREA_SQM + AGE + 
                   PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE +
                   PROX_URA_GROWTH_AREA + PROX_MRT  + PROX_PARK + 
                   PROX_PRIMARY_SCH + PROX_SHOPPING_MALL    + PROX_BUS_STOP + 
                   NO_Of_UNITS + FAMILY_FRIENDLY + FREEHOLD,
                 data=condo_resale.sf)
ols_regress(condo.mlr1)
```

### Preparing Publication Quality Table: gtsummary method

The [**gtsummary**](https://www.danieldsjoberg.com/gtsummary/) package provides an elegant and flexible way to create publication-ready summary tables in R.

```{r}
tbl_regression(condo.mlr1, intercept = TRUE)
```

```{r}
tbl_regression(condo.mlr1, 
               intercept = TRUE) %>% 
  add_glance_source_note(
    label = list(sigma ~ "\U03C3"),
    include = c(r.squared, adj.r.squared, 
                AIC, statistic,
                p.value, sigma))
```

### Checking for multicolinearity

Using `olsrr()` function, which provides a collection of useful methods for building better multiple linear regression models:

-   comprehensive regression output

-   residual diagnostics

-   measures of influence

-   heteroskedasticity tests

-   collinearity diagnostics

-   model fit assessment

-   variable contribution assessment

-   variable selection procedures

The [`ols_vif_tol()`](https://olsrr.rsquaredacademy.com/reference/ols_coll_diag.html) of **olsrr** package is used to test if there are sign of multicollinearity.

```{r}
ols_vif_tol(condo.mlr1)
```

### Test for Non-Linearity

Test the assumption that linearity and additivity of the relationship between dependent and independent variables. The [`ols_plot_resid_fit()`](https://olsrr.rsquaredacademy.com/reference/ols_plot_resid_fit.html) of **olsrr** package is used to perform linearity assumption test.

```{r}
ols_plot_resid_fit(condo.mlr1)
```

The figure above reveals that most of the data points are scattered around the line 0, hence we can conclude that the relationships between the dependent variable and independent variables are linear.

### Test for Normality Assumption

Uses [`ols_plot_resid_hist()`](https://olsrr.rsquaredacademy.com/reference/ols_plot_resid_hist.html) of *olsrr* package to perform normality assumption test.

```{r}
ols_plot_resid_hist(condo.mlr1)
```

The figure reveals that the residual of the multiple linear regression model (i.e. condo.mlr1) is resemble normal distribution.

If you prefer formal statistical test methods, the [`ols_test_normality()`](https://olsrr.rsquaredacademy.com/reference/ols_test_normality.html) of **olsrr** package can be used as shown in the code chun below.

```{r}
ols_test_normality(condo.mlr1)
```

The summary table above reveals that the p-values of the four tests are way smaller than the alpha value of 0.05. Hence we will reject the null hypothesis and infer that there is statistical evidence that the residual are not normally distributed.

### Testing for Spatial Autocorrelation

The hedonic model we try to build are using geographically referenced attributes, hence it is also important for us to visual the residual of the hedonic pricing model.

In order to perform spatial autocorrelation test, we need to convert *condo_resale.sf* from sf data frame into a **SpatialPointsDataFrame**.

First, we will export the residual of the hedonic pricing model and save it as a data frame.

```{r}
mlr.output <- as.data.frame(condo.mlr1$residuals)
```

We will join the newly created data frame with *condo_resale.sf* object.

```{r}
condo_resale.res.sf <- cbind(condo_resale.sf, 
                        condo.mlr1$residuals) %>%
rename(`MLR_RES` = `condo.mlr1.residuals`)
```

We will convert *condo_resale.res.sf* from simple feature object into a SpatialPointsDataFrame because spdep package can only process sp conformed spatial data objects.

```{r}
condo_resale.sp <- as_Spatial(condo_resale.res.sf)
condo_resale.sp
```

We will use **tmap** package to display the distribution of the residuals on an interactive map.

```{r}
tmap_mode("view")
tm_shape(mpsz_svy21)+
  tmap_options(check.and.fix = TRUE) +
  tm_polygons(alpha = 0.4) +
tm_shape(condo_resale.res.sf) +  
  tm_dots(col = "MLR_RES",
          alpha = 0.6,
          style="quantile") +
  tm_view(set.zoom.limits = c(11,14))
```

The figure above reveal that there is sign of spatial autocorrelation.

To proof that our observation is indeed true, the Moran\'s I test will be performed

First, we will compute the distance-based weight matrix by using [`dnearneigh()`](https://r-spatial.github.io/spdep/reference/dnearneigh.html) function of **spdep**.

```{r}
tmap_mode("plot")
nb <- dnearneigh(coordinates(condo_resale.sp), 0, 1500, longlat = FALSE)
summary(nb)
```

Next, [`nb2listw()`](https://r-spatial.github.io/spdep/reference/nb2listw.html) of **spdep** packge will be used to convert the output neighbours lists (i.e. nb) into a spatial weights.

`lm.morantest()` of **spdep** package will be used to perform Moran\'s I test for residual spatial autocorrelation

```{r}
nb_lw <- nb2listw(nb, style = 'W')
summary(nb_lw)
lm.morantest(condo.mlr1, nb_lw)
```

# Building Hedonic Pricing Models using GWmodel

## Fixed Bandwidth GWR Model

### Computing Fixed Bandwidth

In the code chunk below `bw.gwr()` of GWModel package is used to determine the optimal fixed bandwidth to use in the model. Notice that the argument ***adaptive*** is set to **FALSE** indicates that we are interested to compute the fixed bandwidth.

There are two possible approaches can be uused to determine the stopping rule, they are:

-   CV cross-validation approach and AIC corrected (AICc) approach.

-   We define the stopping rule using ***approach*** argeement.

```{r}
bw.fixed <- bw.gwr(formula = SELLING_PRICE ~ AREA_SQM + AGE + PROX_CBD + 
                     PROX_CHILDCARE + PROX_ELDERLYCARE  + PROX_URA_GROWTH_AREA + 
                     PROX_MRT   + PROX_PARK + PROX_PRIMARY_SCH + 
                     PROX_SHOPPING_MALL + PROX_BUS_STOP + NO_Of_UNITS + 
                     FAMILY_FRIENDLY + FREEHOLD, 
                   data=condo_resale.sp, 
                   approach="CV", 
                   kernel="gaussian", 
                   adaptive=FALSE, 
                   longlat=FALSE)
```

### GWModel method - fixed bandwith

```{r}
gwr.fixed <- gwr.basic(formula = SELLING_PRICE ~ AREA_SQM + AGE + PROX_CBD + 
                         PROX_CHILDCARE + PROX_ELDERLYCARE  + PROX_URA_GROWTH_AREA + 
                         PROX_MRT   + PROX_PARK + PROX_PRIMARY_SCH + 
                         PROX_SHOPPING_MALL + PROX_BUS_STOP + NO_Of_UNITS + 
                         FAMILY_FRIENDLY + FREEHOLD, 
                       data=condo_resale.sp, 
                       bw=bw.fixed, 
                       kernel = 'gaussian', 
                       longlat = FALSE)

gwr.fixed
```

## Adaptive Bandwidth GWR Model

### Computing the adaptive bandwidth

Similar to the earlier section, we will first use *bw.ger()* to determine the recommended data point to use.

The code chunk used look very similar to the one used to compute the fixed bandwidth except the ***adaptive*** argument has changed to **TRUE**.

```{r}
bw.adaptive <- bw.gwr(formula = SELLING_PRICE ~ AREA_SQM + AGE  + 
                        PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE    + 
                        PROX_URA_GROWTH_AREA + PROX_MRT + PROX_PARK + 
                        PROX_PRIMARY_SCH + PROX_SHOPPING_MALL   + PROX_BUS_STOP + 
                        NO_Of_UNITS + FAMILY_FRIENDLY + FREEHOLD, 
                      data=condo_resale.sp, 
                      approach="CV", 
                      kernel="gaussian", 
                      adaptive=TRUE, 
                      longlat=FALSE)
bw.adaptive
```


### Constructing the adaptive bandwidth gwr model

```{r}
gwr.adaptive <- gwr.basic(formula = SELLING_PRICE ~ AREA_SQM + AGE + 
                            PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE + 
                            PROX_URA_GROWTH_AREA + PROX_MRT + PROX_PARK + 
                            PROX_PRIMARY_SCH + PROX_SHOPPING_MALL + PROX_BUS_STOP + 
                            NO_Of_UNITS + FAMILY_FRIENDLY + FREEHOLD, 
                          data=condo_resale.sp, bw=bw.adaptive, 
                          kernel = 'gaussian', 
                          adaptive=TRUE, 
                          longlat = FALSE)

gwr.adaptive
```

### Visualizing GWR Output

In addition to regression residuals, the output feature class table includes fields for observed and predicted y values, condition number (cond), Local R2, residuals, and explanatory variable coefficients and standard errors:

-   **Condition Number:** this diagnostic evaluates local collinearity. In the presence of strong local collinearity, results become unstable. Results associated with condition numbers larger than 30, may be unreliable.

-   **Local R2:** these values range between 0.0 and 1.0 and indicate how well the local regression model fits observed y values. Very low values indicate the local model is performing poorly. Mapping the Local R2 values to see where GWR predicts well and where it predicts poorly may provide clues about important variables that may be missing from the regression model.

-   **Predicted:** these are the estimated (or fitted) y values 3. computed by GWR.

-   **Residuals:** to obtain the residual values, the fitted y values are subtracted from the observed y values. Standardized residuals have a mean of zero and a standard deviation of 1. A cold-to-hot rendered map of standardized residuals can be produce by using these values.

-   **Coefficient Standard Error:** these values measure the reliability of each coefficient estimate. Confidence in those estimates are higher when standard errors are small in relation to the actual coefficient values. Large standard errors may indicate problems with local collinearity.

### Converting SDF into sf data.frame

To visualise the fields in **SDF**, we need to first covert it into **sf** data.frame by using the code chunk below.

```{r}
condo_resale.sf.adaptive <- st_as_sf(gwr.adaptive$SDF) %>%
  st_transform(crs=3414)

condo_resale.sf.adaptive.svy21 <- st_transform(condo_resale.sf.adaptive, 3414)

gwr.adaptive.output <- as.data.frame(gwr.adaptive$SDF)
condo_resale.sf.adaptive <- cbind(condo_resale.res.sf, as.matrix(gwr.adaptive.output))
```

Using `glimpse()` to display the content of *condo_resale.sf.adaptive* sf data frame.

```{r}
glimpse(condo_resale.sf.adaptive)
```

```{r}
summary(gwr.adaptive$SDF$yhat)
```

Using `summary()`, we are able to determine the maximum value of 13887901, and its median value is 1385528.

## Visualizing local R2

```{r}
tmap_mode("view")
tm_shape(mpsz_svy21)+
  tm_polygons(alpha = 0.1) +
tm_shape(condo_resale.sf.adaptive) +  
  tm_dots(col = "Local_R2",
          border.col = "gray60",
          border.lwd = 1) +
  tm_view(set.zoom.limits = c(11,14))
```

```{r}
tmap_mode("plot")
```

### By URA Plannign Region

```{r}
tm_shape(mpsz_svy21[mpsz_svy21$REGION_N=="CENTRAL REGION", ])+
  tm_polygons()+
tm_shape(condo_resale.sf.adaptive) + 
  tm_bubbles(col = "Local_R2",
           size = 0.15,
           border.col = "gray60",
           border.lwd = 1)
```
