---
title: "Chapter 5. Global and Local Measures of Spatial Autocorrelation"
editor: visual
---

# Overview

We will be looking on how to delineate homogeneous region by using geographically referenced multivariate data. Looking at the two major analysis, hierarchical cluster analysis and spatially constrained cluster analysis.

# The Study

We are interested to delineate Shan State, Myanmar into homogeneous regions by using multiple Information and Communication technology (ICT) measures, namely: Radio, Television, Land line phone, Mobile phone, Computer, and Internet at home.

# Initialization

Initialize the following packages using pacman:

```{r}
pacman::p_load(rgdal, spdep, tmap, sf, 
               ggpubr, cluster, factoextra, NbClust,
               heatmaply, corrplot, psych, tidyverse)
```

The R packages needed for this exercise are as follows:

-   Spatial data handling

    -   **sf**, **rgdal** and **spdep**

-   Attribute data handling

    -   **tidyverse**, especially **readr**, **ggplot2** and **dplyr**

-   Choropleth mapping

    -   **tmap**

-   Multivariate data visualisation and analysis

    -   **coorplot**, **ggpubr**, and **heatmaply**

-   Cluster analysis

    -   **cluster**

    -   **ClustGeo**

# Data Acquisition and Import

## Data Acquisition

-   Myanmar Township Boundary Data (i.e. *myanmar_township_boundaries*) : This is a GIS data in ESRI shapefile format. It consists of township boundary information of Myanmar. The spatial data are captured in polygon features.

-   *Shan-ICT.csv*: This is an extract of [**The 2014 Myanmar Population and Housing Census Myanmar**](https://myanmar.unfpa.org/en/publications/2014-population-and-housing-census-myanmar-data-sheet) at the township level

## Importing Data

### Geospatial Data

Using st_read() function of sf package to import the relevant shapefile into R.

```{r}
shan_sf <- st_read(dsn = "data/geospatial", 
                   layer = "myanmar_township_boundaries") %>%
  filter(ST %in% c("Shan (East)", "Shan (North)", "Shan (South)"))
```

The Myanmar Township Boundary GIS data is in ESRI shapefile format. It will be imported into R environment by using the [*st_read()*](https://www.rdocumentation.org/packages/sf/versions/0.7-2/topics/st_read) function of **sf**.

### Aspatial Data

The csv file will be import using read_csv function of readr package.

```{r}
ict <- read_csv ("data/aspatial/Shan-ICT.csv")
```

The imported InfoComm variables are extracted from **The 2014 Myanmar Population and Housing Census Myanmar**. The attribute data set is called *ict*. It is saved in R's \* tibble data.frame\* format.

```{r}
summary(ict)
```

## Derive new variables using dplyr package

The unit of measurement of the values are number of household. Using these values directly will be bias by the underlying total number of households. In general, the townships with relatively higher total number of households will also have higher number of households owning radio, TV, etc.

We will derive the penetration rate of each ICT variable by using the following code chunk.

```{r}
ict_derived <- ict %>%
  mutate(`RADIO_PR` = `Radio`/`Total households`*1000) %>%
  mutate(`TV_PR` = `Television`/`Total households`*1000) %>%
  mutate(`LLPHONE_PR` = `Land line phone`/`Total households`*1000) %>%
  mutate(`MPHONE_PR` = `Mobile phone`/`Total households`*1000) %>%
  mutate(`COMPUTER_PR` = `Computer`/`Total households`*1000) %>%
  mutate(`INTERNET_PR` = `Internet at home`/`Total households`*1000) %>%
  rename(`DT_PCODE` =`District Pcode`,`DT`=`District Name`,
         `TS_PCODE`=`Township Pcode`, `TS`=`Township Name`,
         `TT_HOUSEHOLDS`=`Total households`,
         `RADIO`=`Radio`, `TV`=`Television`, 
         `LLPHONE`=`Land line phone`, `MPHONE`=`Mobile phone`,
         `COMPUTER`=`Computer`, `INTERNET`=`Internet at home`) 
```

```{r}
summary(ict_derived)
```

# Exploratory Data Analysis (EDA)

## EDA using statistical graphics

Plot the distribution of the variables (i.e. Number of households with radio) by using appropriate Exploratory Data Analysis (EDA) as shown in the code chunk below.

Histogram is useful to identify the overall distribution of the data values (i.e. left skew, right skew or normal distribution)

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

We will next discover the outliers using boxplot.

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_boxplot(color="black", 
               fill="light blue")
```

Plot the distribution of the newly derived variables (i.e. Radio penetration rate) by using the code chunk below.

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_boxplot(color="black", 
               fill="light blue")
```

Looking at the histogram, the newly derived variables have a normal distribution. Comparing the the boxplot, there's lesser outlier for the newly derived variables.

Let's take a look at multiple histograms are plotted to reveal the distribution of the selected variables in the *ict_derived* data.frame.

```{r}
radio <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

tv <- ggplot(data=ict_derived, 
             aes(x= `TV_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

llphone <- ggplot(data=ict_derived, 
             aes(x= `LLPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

mphone <- ggplot(data=ict_derived, 
             aes(x= `MPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

computer <- ggplot(data=ict_derived, 
             aes(x= `COMPUTER_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

internet <- ggplot(data=ict_derived, 
             aes(x= `INTERNET_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

The [*ggarange()*](https://rpkgs.datanovia.com/ggpubr/reference/ggarrange.html) function of [**ggpubr**](https://rpkgs.datanovia.com/ggpubr/) package is used to group these histograms together.

```{r}
ggarrange(radio, tv, llphone, mphone, computer, internet, 
          ncol = 3, 
          nrow = 2)
```

## Choropleth Map

### Joining geospatial data with aspatial data

```{r}
shan_sf <- left_join(shan_sf, 
                     ict_derived, 
                     by=c("TS_PCODE"="TS_PCODE"))
```

The message above shows that TS_CODE field is the common field used to perform the left-join. It is important to note that there is no new output data been created. Instead, the data fields from ict_derived data frame are now updated into the data frame of shan_sf.

### Plotting the Choropleth Map

To have a quick look at the distribution of Radio penetration rate of Shan State at township level, a choropleth map will be prepared.

```{r}
qtm(shan_sf, "RADIO_PR")
```

In order to reveal the distribution shown in the choropleth map above are bias to the underlying total number of households at the townships, we will create two choropleth maps, one for the total number of households (i.e. TT_HOUSEHOLDS.map) and one for the total number of household with Radio (RADIO.map) by using the code chunk below.

```{r}
TT_HOUSEHOLDS.map <- tm_shape(shan_sf) + 
  tm_fill(col = "TT_HOUSEHOLDS",
          n = 5,
          style = "jenks", 
          title = "Total households") + 
  tm_borders(alpha = 0.5) 

RADIO.map <- tm_shape(shan_sf) + 
  tm_fill(col = "RADIO",
          n = 5,
          style = "jenks",
          title = "Number Radio ") + 
  tm_borders(alpha = 0.5) 

tmap_arrange(TT_HOUSEHOLDS.map, RADIO.map,
             asp=NA, ncol=2)
```

Notice that the choropleth maps above clearly show that townships with relatively larger number ot households are also showing relatively higher number of radio ownership.

Now let us plot the choropleth maps showing the dsitribution of total number of households and Radio penetration rate by using the code chunk below.

```{r}
tmap_mode("view")
tm_shape(shan_sf) +
    tm_polygons(c("TT_HOUSEHOLDS", "RADIO_PR"),
                style="jenks",title = c("Total households","Radio Penetration")) +
    tm_facets(sync = TRUE, ncol = 2) +
  tm_legend(legend.position = c("right", "bottom"))+
  tm_layout(outer.margins=0, asp=0)
```

The lesser the household the higher the radio penetration in some areas.

# Correlation Analysis

Before we perform cluster analysis, it is important for us to ensure that the cluster variables are not highly correlated.

In this section, you will learn how to use [*corrplot.mixed()*](https://cran.r-project.org/web/packages/corrplot/corrplot.pdf) function of [**corrplot**](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html) package to visualise and analyse the correlation of the input variables.

```{r}
cluster_vars.cor = cor(ict_derived[,12:17])
corrplot.mixed(cluster_vars.cor,
         lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

The correlation plot above shows that COMPUTER_PR and INTERNET_PR are highly correlated. This suggest that only one of them should be used in the cluster analysis instead of both.

# Hierarchy Cluster Analysis

## Extrating clustering variables

```{r}
cluster_vars <- shan_sf %>%
  st_set_geometry(NULL) %>%
  select("TS.x", "RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR", "COMPUTER_PR")
head(cluster_vars,10)
```

Notice that the final clustering variables list does not include variable INTERNET_PR because it is highly correlated with variable COMPUTER_PR.

```{r}
row.names(cluster_vars) <- cluster_vars$"TS.x"
head(cluster_vars,10)
```

Notice that the row number has been replaced into the township name.

Now, we will delete the TS.x field by using the code chunk below.

```{r}
shan_ict <- select(cluster_vars, c(2:6))
head(shan_ict, 10)
```

## Data Standardization

It is not unusual their values range are different. In order to avoid the cluster analysis result is baised to clustering variables with large values, it is useful to standardise the input variables before performing cluster analysis.

### Min-Max standardization

In the code chunk below, *normalize()* of [*heatmaply*](https://cran.r-project.org/web/packages/heatmaply/) package is used to stadardisation the clustering variables by using Min-Max method. The *summary()* is then used to display the summary statistics of the standardized clustering variables.

```{r}
shan_ict.std <- normalize(shan_ict)
summary(shan_ict.std)
```

Using Min-Max standardization changes the clustering variable to 0-1.

### Z-score standardization

Z-score standardization can be performed easily by using scale() of Base R. The code chunk below will be used to standardization the clustering variables by using Z-score method.

***Warning: Z-score standardization method should only be used if we would assume all variables come from some normal distribution.***

```{r}
shan_ict.z <- scale(shan_ict)
describe(shan_ict.z)
```

Notice the mean and standard deviation of the Z-score standardized clustering variables are 0 and 1 respectively.

## Visualizing the standardized clustering variables

```{r}
r <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data=shan_ict_s_df, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data=shan_ict_z_df, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

Notice that the overall distribution of the clustering variables will change after the data standardization. Hence, it is advisible **NOT** to perform data standardization if the values range of the clustering variables are not very large.

## Computing proximity matrix

In R, many packages provide functions to calculate distance matrix. We will compute the proximity matrix by using [*dist()*](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/dist.html) of R.

*dist()* supports six distance proximity calculations, they are: euclidean, maximum, manhattan, canberra, binary and minkowski. The default is *euclidean* proximity matrix.

```{r}
proxmat <- dist(shan_ict, method = 'euclidean')
proxmat
```

## Computing hierarchical clustering

[hclust()](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/hclust.html "hclust()") employed agglomeration method to compute the cluster. Eight clustering algorithms are supported, they are: ward.D, ward.D2, single, complete, average(UPGMA), mcquitty(WPGMA), median(WPGMC) and centroid(UPGMC).

The code chunk below performs hierarchical cluster analysis using ward.D method. The hierarchical clustering output is stored in an object of class **hclust** which describes the tree produced by the clustering process.

```{r}
hclust_ward <- hclust(proxmat, method = 'ward.D')
plot(hclust_ward, cex = 0.6)

```

## Selecting the optimal clustering algorithm

One of the challenge in performing hierarchical clustering is to identify stronger clustering structures. The issue can be solved by using use [*agnes()*](https://www.rdocumentation.org/packages/cluster/versions/2.1.0/topics/agnes) function of [**cluster**](https://cran.r-project.org/web/packages/cluster/) package. It functions like *hclus()*, however, with the *agnes()* function you can also get the agglomerative coefficient, which measures the amount of clustering structure found (values closer to 1 suggest strong clustering structure).

```{r}
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")

ac <- function(x) {
  agnes(shan_ict, method = x)$ac
}

map_dbl(m, ac)
```

With reference to the output above, we can see that Ward's method provides the strongest clustering structure among the four methods assessed. Hence, in the subsequent analysis, only Ward's method will be used.

## Determining Optimal Clusters

### Gap Statistic Method

The [**gap statistic**](http://www.web.stanford.edu/~hastie/Papers/gap.pdf) compares the total within intra-cluster variation for different values of k with their expected values under null reference distribution of the data. The estimate of the optimal clusters will be value that maximize the gap statistic (i.e., that yields the largest gap statistic). This means that the clustering structure is far away from the random uniform distribution of points.

To compute the gap statistic, [*clusGap()*](https://www.rdocumentation.org/packages/cluster/versions/2.1.0/topics/clusGap) of [**cluster**](https://cran.r-project.org/web/packages/cluster/) package will be used.

```{r}
set.seed(12345)
gap_stat <- clusGap(shan_ict, 
                    FUN = hcut, 
                    nstart = 25, 
                    K.max = 10, 
                    B = 50)
# Print the result
print(gap_stat, method = "firstmax")
```

Also note that the [*hcut*](https://rpkgs.datanovia.com/factoextra/reference/hcut.html) function used is from [**factoextra**](https://rpkgs.datanovia.com/factoextra/) package.

Next, we can visualise the plot by using [*fviz_gap_stat()*](https://rpkgs.datanovia.com/factoextra/reference/fviz_nbclust.html) of [**factoextra**](https://rpkgs.datanovia.com/factoextra/) package.

```{r}
fviz_gap_stat(gap_stat)
```

With reference to the gap statistic graph above, the recommended number of cluster to retain is 1. However, it is not logical to retain only one cluster. By examine the gap statistic graph, the 6-cluster gives the largest gap statistic and should be the next best cluster to pick.

**Note:** In addition to these commonly used approaches, the [NbClust](https://cran.r-project.org/web/packages/NbClust/) package, published by Charrad et al., 2014, provides 30 indices for determining the relevant number of clusters and proposes to users the best clustering scheme from the different results obtained by varying all combinations of number of clusters, distance measures, and clustering methods.

### Interpreting the dendrograms

In the dendrogram displayed above, each leaf corresponds to one observation. As we move up the tree, observations that are similar to each other are combined into branches, which are themselves fused at a higher height.

The height of the fusion, provided on the vertical axis, indicates the (dis)similarity between two observations. The higher the height of the fusion, the less similar the observations are. Note that, conclusions about the proximity of two observations can be drawn only based on the height where branches containing those two observations first are fused. We cannot use the proximity of two observations along the horizontal axis as a criteria of their similarity.

It's also possible to draw the dendrogram with a border around the selected clusters by using [*rect.hclust()*](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/rect.hclust.html) of R stats. The argument *border* is used to specify the border colors for the rectangles.

```{r}
plot(hclust_ward, cex = 0.6)
rect.hclust(hclust_ward, 
            k = 6, 
            border = 2:5)
```

### Visually-driven hierarchical clustering analysis

In this section, we will learn how to perform visually-driven hiearchical clustering analysis by using [*heatmaply*](https://cran.r-project.org/web/packages/heatmaply/) package.

With **heatmaply**, we are able to build both highly interactive cluster heatmap or static cluster heatmap.

We will need transform shan_ict data frame into a data matrix, and plot the the interactive cluster heatmap using `heatmaply()`.

```{r}
shan_ict_mat <- data.matrix(shan_ict)

heatmaply(normalize(shan_ict_mat),
          Colv=NA,
          dist_method = "euclidean",
          hclust_method = "ward.D",
          seriate = "OLO",
          colors = Blues,
          k_row = 6,
          margins = c(NA,200,60,NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main="Geographic Segmentation of Shan State by ICT indicators",
          xlab = "ICT Indicators",
          ylab = "Townships of Shan State"
          )
```

### Mapping the clusters formed

With closed examination of the dendragram above, we have decided to retain six clusters.

[*cutree()*](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/cutree.html) of R Base will be used in the code chunk below to derive a 6-cluster model.

```{r}
groups <- as.factor(cutree(hclust_ward, k=6))
```

The output is called *groups*. It is a *list* object.

In order to visualise the clusters, the *groups* object need to be appended onto *shan_sf* simple feature object.

The code chunk below form the join in three steps:

-   the *groups* list object will be converted into a matrix;

-   *cbind()* is used to append *groups* matrix onto shan_sf to produce an output simple feature object called `shan_sf_cluster`; and

-   *rename* of **dplyr** package is used to rename *as.matrix.groups* field as *CLUSTER*.

```{r}
shan_sf_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER`=`as.matrix.groups.`)
qtm(shan_sf_cluster, "CLUSTER")
```

The choropleth map above reveals the clusters are very fragmented. The is one of the major limitation when non-spatial clustering algorithm such as hierarchical cluster analysis method is used.

# Spatially Constrained Clustering - SKATER approach

## Converting into SpatialPolygonsDataFrame

The following code chunk uses [*as_Spatial()*](https://r-spatial.github.io/sf/reference/coerce-methods.html) of **sf** package to convert *shan_sf* into a SpatialPolygonDataFrame called *shan_sp*.

```{r}
shan_sp <- as_Spatial(shan_sf)
```

## Computing Neighbour List

Next, poly2nd() of spdep package will be used to compute the neighbours list from polygon list.

```{r}
shan.nb <- poly2nb(shan_sp)
summary(shan.nb)
```

We can plot the neighbours list on shan_sp by using the code chunk below. Since we now can plot the community area boundaries as well, we plot this graph on top of the map. The first plot command gives the boundaries. This is followed by the plot of the neighbor list object, with coordinates applied to the original SpatialPolygonDataFrame (Shan state township boundaries) to extract the centroids of the polygons. These are used as the nodes for the graph representation. We also set the color to blue and specify add=TRUE to plot the network on top of the boundaries.

```{r}
plot(shan_sp, 
     border=grey(.5))
plot(shan.nb, 
     coordinates(shan_sp), 
     col="blue", 
     add=TRUE)
```

Note that if you plot the network first and then the boundaries, some of the areas will be clipped. This is because the plotting area is determined by the characteristics of the first plot. In this example, because the boundary map extends further than the graph, we plot it first.

## Computing minimum spanning tree

Next, [*nbcosts()*](https://r-spatial.github.io/spdep/reference/nbcosts.html) of **spdep** package is used to compute the cost of each edge. It is the distance between it nodes. This function compute this distance using a data.frame with observations vector in each node.

We will need to compute the cost of each edge and convert the neighbor list to a list weights object by specifying the just computed lcosts as the weights.

```{r}
lcosts <- nbcosts(shan.nb, shan_ict)

shan.w <- nb2listw(shan.nb, 
                   lcosts, 
                   style="B")
summary(shan.w)
```

Note that we specify the *style* as **B** to make sure the cost values are not row-standardized.

## Computing minimum spanning tree

The minimum spanning tree is computed by mean of the [*mstree()*](https://r-spatial.github.io/spdep/reference/mstree.html) of **spdep** package as shown in the code chunk below.

```{r}
shan.mst <- mstree(shan.w)

class(shan.mst)

dim(shan.mst)
```

Note that the dimension is 54 and not 55. This is because the minimum spanning tree consists on n-1 edges (links) in order to traverse all the nodes.

We can display the content of *shan.mst* by using *head()* as shown in the code chunk below.

The plot method for the MST include a way to show the observation numbers of the nodes in addition to the edge. As before, we plot this together with the township boundaries. We can see how the initial neighbor list is simplified to just one edge connecting each of the nodes, while passing through all the nodes.

```{r}
plot(shan_sp, border=gray(.5))
plot.mst(shan.mst, 
         coordinates(shan_sp), 
         col="blue", 
         cex.lab=0.7, 
         cex.circles=0.005, 
         add=TRUE)
```

## Computing spatially constrained clusters using SKATER method

Compute the spatially constrained cluster using [*skater()*](https://r-spatial.github.io/spdep/reference/skater.html) of **spdep** package.

```{r}
clust6 <- skater(edges = shan.mst[,1:2], 
                 data = shan_ict, 
                 method = "euclidean", 
                 ncuts = 5)
```

The *skater()* takes three mandatory arguments:

1.  the first two columns of the MST matrix (i.e. not the cost),

2.  the data matrix (to update the costs as units are being grouped), and

3.  the number of cuts.

Note: It is set to **one less than the number of clusters**. So, the value specified is **not** the number of clusters, but the number of cuts in the graph, one less than the number of clusters.

```{r}
str(clust6)
```

The most interesting component of this list structure is the groups vector containing the labels of the cluster to which each observation belongs (as before, the label itself is arbitary).

```{r}
ccs6 <- clust6$groups

plot(shan_sp, border=gray(.5))
plot(clust6, 
     coordinates(shan_sp), 
     cex.lab=.7,
     groups.colors=c("red","green","blue", "brown", "pink"),
     cex.circles=0.005, 
     add=TRUE)
```

## Visualizing the clusters in choropleth map

Plot the newly derived clusters by using SKATER method. For easy comparison, it will be better to place both the hierarchical clustering and spatially constrained hierarchical clustering maps next to each other.

```{r}
groups_mat <- as.matrix(clust6$groups)
shan_sf_spatialcluster <- cbind(shan_sf_cluster, as.factor(groups_mat)) %>%
  rename(`SP_CLUSTER`=`as.factor.groups_mat.`)

hclust.map <- qtm(shan_sf_cluster,
                  "CLUSTER") + 
  tm_borders(alpha = 0.5) 

shclust.map <- qtm(shan_sf_spatialcluster,
                   "SP_CLUSTER") + 
  tm_borders(alpha = 0.5) 

tmap_arrange(hclust.map, shclust.map,
             asp=NA, ncol=2)
```
